---
title: "Fancy: Hybrid Network Inference for Microbiome Co-Abundance Analysis"
author: "Wanxin Lai"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Fancy workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  crop     = if (requireNamespace("magick", quietly = TRUE)) "magick" else NULL
)
```

```{css custom-styles, echo=FALSE}
h1, h2, h3, h4 { color: #404040; }
h1 { border-bottom: 1px #404040 solid; }
span.caption-title { color: #404040; }
table {
  margin-left: auto !important;
  margin-right: auto !important;
  float: none !important;
}
.formula-box {
  padding: 15px 20px;
  margin: 20px 0;
  text-align: center;
  font-size: 1.1em;
}
.formula-box .MathJax_Display {
  margin: 0.5em 0 !important;
}
```

# Introduction

**Fancy** infers microbial co-abundance networks using a hybrid scoring
framework that combines kNN-based mutual information (MI), MRNET edge
selection, and distance correlation (dCor) within a bootstrap resampling
loop. Each bootstrap iteration computes an MI matrix, applies MRNET for
sparse edge selection, and independently computes pairwise dCor. The
final hybrid score ranks edges by combining MRNET edge frequency with
dCor stability across bootstraps.

This hybrid approach captures **nonlinear** associations between MAGs,
including threshold effects, competitive exclusion, and other ecological
interactions that linear correlation measures miss.

```{r load-library, message=FALSE}
library(Fancy)
```

# Example Data

Fancy ships with a dataset subset from a published cattle rumen
metagenome study. The data contains **100 MAGs across 321 samples**
with CLR-transformed abundances pre-computed on the full 2178-MAG
compositional reference:

- **MAGs 1-2**: Bulleidia vs AC2028 (both Bacillota), showing
  threshold competitive exclusion
- **MAGs 3-4**: RUG023 (Spirochaetota) vs Cryptobacteroides
  (Bacteroidota), showing an L-shaped nonlinear relationship
- **MAGs 5-100**: high-quality MAGs spanning 9 phyla

```{r load-data}
data(fancy_tiny_clr)
data(fancy_tiny_counts)
data(fancy_tiny_coverage)
data(fancy_tiny_taxonomy)
data(fancy_tiny_metadata)

dim(fancy_tiny_clr)    # MAGs x samples (CLR-transformed)
fancy_tiny_clr[1:5, 1:5]
```

The taxonomy spans 9 phyla:

```{r phyla-barplot, fig.width=7, fig.height=4}
phyla_counts <- sort(table(fancy_tiny_taxonomy$Phyla), decreasing = TRUE)
par(mar = c(5, 10, 2, 1))
barplot(phyla_counts, horiz = TRUE, las = 1,
        col = phyla_palette(names(phyla_counts))[names(phyla_counts)],
        xlab = "Number of MAGs", main = "Phyla Distribution")
```

## Coverage distribution

Coverage values reflect real genome mapping fractions. Most MAGs in this
subset have high coverage, as they were selected for genome quality
(coverage >= 0.5 in >= 200 samples).

```{r coverage-hist, fig.width=7, fig.height=4}
cov_vals <- unlist(fancy_tiny_coverage)
hist(cov_vals, breaks = 60, main = "Genome Coverage Distribution",
     xlab = "Covered fraction", col = "steelblue", border = "white")
abline(v = 0.3, col = "red", lwd = 2, lty = 2)
text(0.32, par("usr")[4] * 0.85, "min_coverage = 0.3",
     col = "red", cex = 0.8, pos = 4)
cat("Fraction with coverage >= 0.3:", round(mean(cov_vals >= 0.3, na.rm = TRUE), 4), "\n")
```

## Metadata

The metadata links each sample to methane emissions (CH4) and cattle
breed.

```{r metadata, fig.width=7, fig.height=4}
head(fancy_tiny_metadata)
hist(fancy_tiny_metadata$CH4, breaks = 15, main = "CH4 Emissions",
     xlab = "CH4 (g/day)", col = "grey80", border = "white")
```

# Preparing the Network Input

The shipped CLR matrix (`fancy_tiny_clr`) was computed on the full
2178-MAG compositional reference, preserving the proper geometric mean.
Network functions in Fancy expect **samples as rows, MAGs as columns**,
so we transpose:

```{r transpose}
net_input <- t(fancy_tiny_clr)
dim(net_input)  # samples x MAGs
net_input[1:5, 1:5]
```

**Note:** If you are working with your own raw count data, run
`filter_mags()` then `clr_normalize()` before transposing. The example
data ships pre-CLR to avoid recomputing CLR on a reduced MAG set, which
would distort the geometric mean.

# Why a Hybrid Scoring Framework?

Fancy combines two complementary approaches in a bootstrap framework:

1. **Mutual information + MRNET**: kNN-based MI captures *all* types
   of statistical dependencies (linear and nonlinear). MRNET then
   sparsifies the MI network by removing indirect edges, retaining only
   the most informative direct associations. The fraction of bootstrap
   iterations in which MRNET selects an edge gives the **EdgeFrequency**.

2. **Distance correlation (dCor)**: Computed independently in each
   bootstrap iteration, dCor provides a second nonlinear dependency
   measure. The ratio of mean dCor to its standard deviation across
   bootstraps gives a **Stability** score, meaning edges with
   consistently high dCor are more trustworthy.

The final **HybridScore** ranks edges by combining both signals.
Fancy offers two scoring modes:

## Multiplicative scoring (default)

<div class="formula-box">
$$\text{HybridScore} \;=\; \text{EdgeFrequency}^{\,w_1} \;\times\; \text{Stability.dcor.scaled}^{\,w_2}$$
</div>

$w_1$ controls the weight on MRNET edge frequency and $w_2$ controls
the weight on dCor stability (default $w_1 = 0.3$, $w_2 = 0.7$).
Edges need support from **both** components: a near-zero value in
either pulls the score toward zero.

## Additive scoring

<div class="formula-box">
$$\text{HybridScore} \;=\; w_1 \times \text{EdgeFrequency} \;+\; w_2 \times \text{Stability.dcor.scaled}$$
</div>

Set `score_type = "additive"` in `fancy()` to use this mode. This
allows edges with strong dCor but weak MRNET frequency (or vice versa)
to retain a meaningful score — useful when users want purely
dCor-discovered edges to carry weight.

## dCor rescue (multiplicative mode)

For multiplicative scoring, a `dcor_rescue_percentile` option is
available (e.g. `dcor_rescue_percentile = 0.6`). Edges whose
`Stability.dcor.scaled` exceeds the given percentile receive a minimum
`EdgeFrequency` floor (the 30th percentile of non-zero EdgeFrequency
values). This prevents MRNET from suppressing edges that dCor finds
reliably — important for users who believe some associations are only
detectable by dCor and not MI/MRNET.

This hybrid approach captures nonlinear ecological interactions that
Pearson or Spearman correlation would miss.

# Running the Pipeline

`fancy()` wraps the full pipeline: bootstrap network inference, edge
scoring, and thresholding.

| Parameter | Default | Suggested | Description |
|-----------|---------|-----------|-------------|
| `n_bootstrap` | 100 | 100 to 200 | Number of bootstrap iterations |
| `k` | 5 | 3 to 7 | kNN parameter for MI estimation |
| `find_k` | FALSE | TRUE for exploratory runs | Auto-select k via elbow method |
| `cpus` | 4 | 2 to 8 | Number of parallel workers |
| `use_mrnet` | TRUE | TRUE | Apply MRNET for sparse edge selection |
| `w1` | 0.3 | 0.2 to 0.4 | Weight for edge frequency |
| `w2` | 0.7 | 0.6 to 0.8 | Weight for dCor stability |
| `threshold_method` | "quantile" | "quantile" or "top_n" | How to threshold edges |
| `threshold_value` | 0.7 | 0.7 to 0.9 | Quantile cutoff (top 30 percent at 0.7) |

We run **100 bootstrap iterations** here for meaningful hybrid scores.
We set `cpus = 6L` for parallel processing and use the default 70th
percentile threshold. Because this dataset contains high-quality MAGs
subset from real data (100 MAGs, 321 samples), we use **equal weights**
(`w1 = w2 = 0.5`) so that MRNET edge frequency and dCor stability
contribute equally to the hybrid score.

```{r run-fancy, message=FALSE}
result <- fancy(
  net_input,
  n_bootstrap     = 100L,
  k               = 3L,
  cpus            = 6L,
  w1              = 0.5,
  w2              = 0.5,
  threshold_method = "quantile",
  threshold_value  = 0.7,
  verbose          = TRUE
)
```

# Inspecting Results

The result is a list with class `"fancy"`:

```{r structure}
str(result, max.level = 1)
cat("\nRetained edges:", nrow(result$edges), "\n")
cat("Total scored edges:", nrow(result$all_edges), "\n")
```

With 100 MAGs there are `choose(100, 2)` = 4,950 possible pairwise
edges. After bootstrap filtering (minimum EdgeFrequency and dCor
stability), `r nrow(result$all_edges)` edges are scored. The 70th
percentile threshold retains `r nrow(result$edges)` of these.

Top 5 edges by hybrid score:

```{r top-edges}
top5 <- head(result$edges[order(-result$edges$HybridScore), ], 5)
knitr::kable(as.data.frame(top5), digits = 3, row.names = FALSE)
```

## Diagnostic Table for Known Nonlinear Pairs

The `all_edges_unfiltered` component contains all
`r nrow(result$all_edges_unfiltered)` pairs with their intermediate
metrics, including pairs that were dropped by the hard pre-filter.
This is useful for inspecting known nonlinear pairs:

```{r diagnostic-table}
## Known nonlinear pairs (MAGs 1-4)
pairs <- list(c("MAG_001", "MAG_002"), c("MAG_003", "MAG_004"))
diag_rows <- lapply(pairs, function(p) {
  uf <- result$all_edges_unfiltered
  idx <- which((uf$source == p[1] & uf$target == p[2]) |
               (uf$source == p[2] & uf$target == p[1]))
  if (length(idx)) as.data.frame(uf[idx[1], ])
})
diag_df <- do.call(rbind, diag_rows)
knitr::kable(
  diag_df[, c("source", "target", "EdgeFrequency", "mean.dcor", "sd.dcor",
              "Stability.dcor", "Stability.dcor.scaled", "HybridScore")],
  digits = 3, row.names = FALSE,
  caption = "Intermediate metrics for known nonlinear MAG pairs"
)
```

The table shows where each pair sits: **EdgeFrequency** (how often
MRNET selected the edge across bootstraps), **mean.dcor** and
**sd.dcor** (dCor mean and variability), **Stability.dcor** (mean/sd
ratio), and the final **HybridScore**. If a pair has EdgeFrequency
below 0.3, it was dropped from `all_edges` by the hard pre-filter
but remains visible here.

```{r helper-functions, include=FALSE}
## Helper: look up hybrid score for a MAG pair
get_hybrid_score <- function(all_edges, mag1, mag2) {
  idx <- which(
    (all_edges$source == mag1 & all_edges$target == mag2) |
    (all_edges$source == mag2 & all_edges$target == mag1)
  )
  if (length(idx) > 0) round(all_edges$HybridScore[idx[1]], 3) else NA
}

## CH4 colour mapping
ch4_vals  <- fancy_tiny_metadata$CH4
ch4_pal   <- colorRampPalette(c("dodgerblue", "gold", "firebrick"))(100)
ch4_idx   <- as.integer(cut(ch4_vals, breaks = 100))
ch4_cols  <- ch4_pal[ch4_idx]
```

# Edges Across Hybrid Score Bands

The scatter plots below show pairs sampled across four hybrid score
bands, annotated with hybrid scores and Pearson r. Note how edges with
high hybrid scores can have weak Pearson r, indicating nonlinear
dependencies that linear measures miss.

```{r score-band-scatter, fig.width=6, fig.height=12.5}
all_scored <- result$all_edges[order(-result$all_edges$HybridScore), ]

## Define five score bands and sample 2 pairs from each
bands <- list(
  c(0.9, Inf),
  c(0.6, 0.7),
  c(0.45, 0.55),
  c(0.3, 0.4),
  c(0.1, 0.2)
)
band_labels <- c("~1.0", "0.6-0.7", "0.45-0.55", "0.3-0.4", "0.1-0.2")

set.seed(42)
show_pairs <- data.frame()
for (b in seq_along(bands)) {
  lo <- bands[[b]][1]
  hi <- bands[[b]][2]
  in_band <- all_scored[all_scored$HybridScore >= lo &
                        all_scored$HybridScore < hi, ]
  if (nrow(in_band) == 0) {
    ## If exact band is empty, take the 2 closest edges to midpoint
    mid <- (lo + hi) / 2
    all_scored$dist_tmp <- abs(all_scored$HybridScore - mid)
    in_band <- head(all_scored[order(all_scored$dist_tmp), ], 2)
    all_scored$dist_tmp <- NULL
  }

  ## For mid bands (0.3-0.55): prefer low sd.dcor + weak Pearson r
  ## For lowest band (0.1-0.2): prefer higher Pearson r to show that
  ## linearly correlated pairs correctly receive low hybrid scores
  if ((hi <= 0.55 && lo >= 0.3) && nrow(in_band) > 2) {
    in_band$abs_r <- vapply(seq_len(nrow(in_band)), function(j) {
      abs(cor(net_input[, in_band$source[j]],
              net_input[, in_band$target[j]]))
    }, numeric(1))
    low_r <- in_band[in_band$abs_r < 0.4, ]
    if (nrow(low_r) >= 2) {
      low_r <- low_r[order(low_r$sd.dcor), ]
      picked <- head(low_r, 2)
    } else {
      in_band <- in_band[order(in_band$sd.dcor), ]
      picked <- head(in_band, 2)
    }
    picked$abs_r <- NULL
  } else if (lo < 0.3 && nrow(in_band) > 2) {
    in_band$abs_r <- vapply(seq_len(nrow(in_band)), function(j) {
      abs(cor(net_input[, in_band$source[j]],
              net_input[, in_band$target[j]]))
    }, numeric(1))
    ## Pick one pair with |r| > 0.5, one with |r| > 0.4
    high_r <- in_band[order(-in_band$abs_r), ]
    picked <- head(high_r[high_r$abs_r > 0.4, ], 2)
    if (nrow(picked) < 2) picked <- head(high_r, 2)
    picked$abs_r <- NULL
  } else {
    n_pick <- min(2, nrow(in_band))
    picked <- in_band[sample(seq_len(nrow(in_band)), n_pick), ]
  }

  picked$band <- band_labels[b]
  show_pairs <- rbind(show_pairs, picked)
}

par(mfrow = c(5, 2), mar = c(4, 4.5, 3, 1))
for (i in seq_len(nrow(show_pairs))) {
  s <- show_pairs$source[i]
  t <- show_pairs$target[i]
  hs <- round(show_pairs$HybridScore[i], 3)
  band <- show_pairs$band[i]

  s_genus <- fancy_tiny_taxonomy[s, "Genus"]
  t_genus <- fancy_tiny_taxonomy[t, "Genus"]

  xi <- net_input[, t]
  yi <- net_input[, s]
  cti <- cor.test(xi, yi)

  plot(xi, yi,
       xlab = paste0(t_genus, " (CLR)"),
       ylab = paste0(s_genus, " (CLR)"),
       main = paste0("Score ", band, ": ", s_genus, " vs ", t_genus),
       pch = 19, col = adjustcolor(ch4_cols, 0.7), cex = 1.1)
  legend("topright", bty = "n", cex = 0.9,
         legend = c(
           paste0("Hybrid = ", hs),
           paste0("r = ", round(cti$estimate, 3),
                  " (p = ", format.pval(cti$p.value, digits = 2), ")")
         ))
  if (i == 1) {
    legend("bottomleft", bty = "n", cex = 0.8, title = "CH4",
           legend = c("Low", "Mid", "High"),
           pch = 19, col = c("dodgerblue", "gold", "firebrick"))
  }
}
```

As the score bands illustrate, the hybrid framework surfaces nonlinear
patterns across a range of strengths. We recommend inspecting scatter
plots for top-ranking edges to confirm the biological context before
interpretation.

# Nonlinear Interactions Fancy Captured

MAGs 1-4 contain nonlinear associations. We plot their
CLR-transformed abundances and annotate with the hybrid scores
from the pipeline.

```{r scatter-pair1, fig.width=4, fig.height=3.5}
# ---- Pair 1: Threshold competitive exclusion (Bulleidia vs AC2028) ----
x1 <- net_input[, "MAG_002"]
y1 <- net_input[, "MAG_001"]
ct1 <- cor.test(x1, y1)
hs1 <- get_hybrid_score(result$all_edges_unfiltered, "MAG_001", "MAG_002")

par(mar = c(4.5, 4.5, 3.5, 1))
plot(x1, y1,
     xlab = "AC2028 (CLR)",
     ylab = "Bulleidia (CLR)",
     main = "MAG_001 vs MAG_002\nThreshold Competitive Exclusion",
     cex.main = 0.85, cex.axis = 0.75, cex.lab = 0.85,
     pch = 19, col = adjustcolor(ch4_cols, 0.8), cex = 0.9)
lo <- loess(y1 ~ x1, span = 0.75)
ox <- order(x1)
lines(x1[ox], predict(lo)[ox], col = "black", lwd = 2.5)

legend("topright", bty = "n", cex = 0.75,
       legend = c(
         paste0("Pearson r = ", round(ct1$estimate, 3),
                " (p = ", format.pval(ct1$p.value, digits = 2), ")"),
         paste0("Hybrid score = ", hs1)
       ))
legend("bottomleft", bty = "n", cex = 0.7, title = "CH4 (g/day)",
       legend = c("Low", "Mid", "High"),
       pch = 19, col = c("dodgerblue", "gold", "firebrick"))
```

```{r scatter-pair2, fig.width=4, fig.height=3.5}
# ---- Pair 2: L-shaped relationship (RUG023 vs Cryptobacteroides) ----
x2 <- net_input[, "MAG_004"]
y2 <- net_input[, "MAG_003"]
ct2 <- cor.test(x2, y2)
hs2 <- get_hybrid_score(result$all_edges_unfiltered, "MAG_003", "MAG_004")

par(mar = c(4.5, 4.5, 3.5, 1))
plot(x2, y2,
     xlab = "Cryptobacteroides (CLR)",
     ylab = "RUG023 (CLR)",
     main = "MAG_003 vs MAG_004\nL-Shaped Nonlinear Relationship",
     cex.main = 0.85, cex.axis = 0.75, cex.lab = 0.85,
     pch = 19, col = adjustcolor(ch4_cols, 0.8), cex = 0.9)

legend("topright", bty = "n", cex = 0.75,
       legend = c(
         paste0("Pearson r = ", round(ct2$estimate, 3),
                " (p = ", format.pval(ct2$p.value, digits = 2), ")"),
         paste0("Hybrid score = ", hs2)
       ))
legend("bottomleft", bty = "n", cex = 0.7, title = "CH4 (g/day)",
       legend = c("Low", "Mid", "High"),
       pch = 19, col = c("dodgerblue", "gold", "firebrick"))
```

Both interactions have weak Pearson correlations, yet the underlying
biological dependencies are clear. Note that the L-shaped pair
(MAG_003 vs MAG_004) is **lost from the final network** because its
EdgeFrequency falls below the hard hybrid cut-off of 0.3 — MRNET does
not consistently select it across bootstraps, even though dCor captures
the pattern reliably. This is a deliberate trade-off: the multiplicative
hybrid formula prioritises edges supported by **both** MI/MRNET and
dCor, favouring robustness over sensitivity. Purely nonlinear
associations that MI misses can still be found by inspecting
`all_edges_unfiltered` sorted by `Stability.dcor.scaled`. Users who
prefer dCor-discovered edges to be retained can use
`score_type = "additive"` (where strong dCor alone carries a meaningful
score) or set `dcor_rescue_percentile` (e.g. 0.6) to give high-stability
edges a minimum EdgeFrequency floor in the multiplicative formula.

# Re-thresholding

`threshold_edges()` re-thresholds the full scored edge list without
re-running the bootstrap. This is useful for exploring different
stringency levels.

| Parameter | Default | Suggested | Description |
|-----------|---------|-----------|-------------|
| `method` | "quantile" | "quantile" or "top_n" | Thresholding strategy |
| `value` | 0.8 | 0.7 to 0.9 for quantile; 10 to 50 for top_n | Cutoff value |

```{r rethreshold}
strict <- threshold_edges(result$all_edges, method = "quantile", value = 0.8)
cat("Edges at 80th percentile:", nrow(strict), "\n")
```

# Score Distribution

The S3 `plot()` method shows the hybrid score distribution with the
threshold cutoff marked.

```{r plot-fancy, fig.width=7, fig.height=5}
plot(result)
```

# Taxonomy Annotation

Merge taxonomy onto edges to see which lineages are co-abundant.

```{r annotate}
edges <- result$edges

# Add source taxonomy
edges <- merge(edges, fancy_tiny_taxonomy[, c("Genus", "Phyla")],
               by.x = "source", by.y = "row.names", all.x = TRUE)
names(edges)[names(edges) == "Genus"] <- "source_Genus"
names(edges)[names(edges) == "Phyla"] <- "source_Phyla"

# Add target taxonomy
edges <- merge(edges, fancy_tiny_taxonomy[, c("Genus", "Phyla")],
               by.x = "target", by.y = "row.names", all.x = TRUE)
names(edges)[names(edges) == "Genus"] <- "target_Genus"
names(edges)[names(edges) == "Phyla"] <- "target_Phyla"

head(edges[order(-edges$HybridScore),
     c("source_Genus", "target_Genus", "source_Phyla", "target_Phyla",
       "HybridScore")], 15)
```

# Network Visualization

If `igraph` is installed, `plot_network()` draws the co-abundance network
with nodes coloured by phylum and edges weighted by HybridScore.
Communities are detected using modularity-based clustering (analogous
to GLay in Cytoscape). Shaded regions group MAGs that co-occur more
frequently with each other than with MAGs outside their cluster.

```{r plot-network, fig.width=9, fig.height=9, eval=requireNamespace("igraph", quietly=TRUE)}
plot_network(result, fancy_tiny_taxonomy, community = TRUE)
```

# Exporting for Cytoscape

`export_cytoscape()` writes node and edge tables as tab-separated files
that Cytoscape can import directly.

| Parameter | Default | Description |
|-----------|---------|-------------|
| `fancy_result` | | Object of class `"fancy"` |
| `taxonomy` | | Data.frame with MAG IDs as rownames, must have `Phyla` column |
| `file_prefix` | "fancy_network" | Path prefix for output TSV files |
| `edges` | "thresholded" | Which edges: `"thresholded"` or `"all"` |

```{r export-build, eval=TRUE}
cyto <- export_cytoscape(
  result,
  fancy_tiny_taxonomy,
  file_prefix = tempfile("fancy")
)
```

The returned list contains the same data.frames that were written to
disk:

```{r export-edges}
head(cyto$edges)
```

```{r export-nodes}
head(cyto$nodes)
```

To import into Cytoscape:

1. **File > Import > Network from File**: select the `*_edges.tsv` file.
   Map `source` to Source Node, `target` to Target Node, and
   `HybridScore` to an edge attribute.
2. **File > Import > Table from File**: select the `*_nodes.tsv` file.
   Map the `id` column to the node key column.
3. Use the `Color` column in the node table to set node fill colour by
   mapping it as a passthrough in the Style panel.
4. **Apps > Community Detection (GLay)** to run GLay community
   clustering on the network.

# Saving Results

To save all pipeline and intermediate objects for later use without
re-running the bootstrap:

```{r save-objects, eval=TRUE}
save(result, net_input,
     fancy_tiny_clr, fancy_tiny_counts, fancy_tiny_coverage,
     fancy_tiny_taxonomy, fancy_tiny_metadata,
     all_scored, show_pairs, diag_df, edges, top5, strict,
     file = "vignette_objects.RData")
```

# Session Info

```{r session-info}
sessionInfo()
```
